---
title: Topic 8 -- A modern approach to hypothesis testing with infer
author: Floris van Ogtrop
format:
  ochre-revealjs: default
  ochre-typst:
    fig-format: png
---

## Announcement

- Practice Skills assessment this week, run in your respective labs. **You must attend your lab to complete the assessment.**
- A reminder that you are allowed to use cheat sheets for the assessment, but you must write or print them yourself
  - maximum 2 sides of A4 paper
  - no electronic devices allowed
- Printing services are available on campus. See [here](https://www.sydney.edu.au/students/student-it/print-scan-copy.html).

# Learning Outcomes

## By the end of this lecture, you can:

1. Understand hypothesis testing as a **comparison to a distribution**
2. Use simulation to **generate that distribution from data**
3. Perform a hypothesis test using the **infer package**
4. Interpret results without relying on strict assumptions



# The Big Idea

## How do we usually test hypotheses?

- Compare our observed result to a **theoretical distribution**
- This distribution is based on assumptions about the data (e.g., normality, equal variances)


## A different idea

> What if we **didn’t assume a distribution at all**?

- Use the data to **simulate the distribution** of the test statistic under the null hypothesis  
- Simulate “no effect” data by randomly shuffling group labels (permutation test)  
- Compare our observed statistic to this simulated distribution to get a p-value


## Example: Sleep data

```{r}
library(tidyverse)
glimpse(sleep)
```

## Step 1 — Plot the data (EDA)

```{r}
sleep %>%
  ggplot(aes(x = group, y = extra)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  theme_minimal()
```


## Step 2 Observed effect

- Create an observed test statistic using the infer workflow
- Take the 'sleep' dataset and pass it into the infer pipeline
- Define the response variable ('extra') and explanatory variable ('group') i.e., we are comparing 'extra' across the two groups
- Compute the statistic of interest: difference in group means
- Specify the order: mean(group 1) - mean(group 2). This determines the direction (sign) of the difference

```{r}
library(infer)
observed <- sleep %>% 
  specify(extra ~ group) %>%
  calculate(stat = "diff in means",           
            order = c(1, 2)) 
observed
```

# Think first (Prediction)

## Before we run the test:

- Does the difference look **big or small**?
- Do you expect:
  - A) Small p-value (< 0.05)
  - B) Moderate (~0.1)
  - C) Large (> 0.2)

> Commit to an answer

## Step 3 — Simulate null distribution (HAT)

- Start with the dataset (sleep) to build a null distribution
- Set the seed for reproducibility
- Specify the response (extra) and grouping variable (group)
- **(H)** State the null hypothesis of no relationship (independence) - Assume the response is unrelated to group (i.e., group labels don’t matter).
- **(A)** Randomly shuffle group labels 1000 times to simulate “no effect” data (*exchangeability under the null hypothesis*)
- **(T)** Calculate the difference in means for each simulated dataset

```{r}
set.seed(123)

null_dist <- sleep %>%
  specify(extra ~ group) %>%
  hypothesise(null = "independence") %>% ## Hypothesis (H)
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in means", order = c(1, 2)) # Test (T)
```


## Step 3 — p-value (P)

```{r}
p_value <- null_dist %>%
  get_p_value(obs_stat = observed, direction = "two-sided")

p_value
```

- we can visualise the null distribution and see where our observed statistic falls, and shade the area corresponding to the p-value

```{r}
null_dist %>%
  visualize() +
  shade_p_value(obs_stat = observed, direction = "two-sided")
```

## Conclusion (C)

> If the observed statistic is rare under “no effect” → evidence against null
> In this case it is not super rare, so we do not have strong evidence against the null hypothesis of no difference in sleep between the two groups.

## Compare to t-test

- The t-test also tests the null hypothesis of no difference between groups, but it relies on assumptions (e.g., normality, equal variances).
- The permutation test we performed does not rely on these assumptions and is based on the actual data.
- The p-value from the permutation test may differ from the t-test, especially if the assumptions of the t-test are violated. In this case, we might find that the t-test gives a different p-value, which could lead to a different conclusion about the evidence against the null hypothesis.

```{r}
t.test(extra ~ group, data = sleep)
```


# Summary

## Key points

- Hypothesis testing = compare to distribution  
- We can simulate that distribution  
- `infer` provides a clean workflow  

simulate → compare → conclude

## Final Thought

> We are not memorising tests  
> We are learning how to **"build" inference**


# Thanks!

This presentation is based on the [SOLES Quarto reveal.js template](https://github.com/usyd-soles-edu/soles-revealjs) and is licensed under a [Creative Commons Attribution 4.0 International License][cc-by].


<!-- Links -->
[cc-by]: http://creativecommons.org/licenses/by/4.0/
